---
title: "cyano_bubble_plots.Rmd"
author: "Dylan Barth"
date: "2026-03-29"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
library(tidyverse)
library(phyloseq)
library(circlize)
library(scales)  
library(shiny)
library(rlang)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/WorkForaging/Academia/Nicole/deca/')
```

## Cyanobacteria Relative Abundance Interactive

Here, since ANCOMBC isn't exactly appropriate for this subset of data, we show average relative abundance at various taxonomic levels across a chosen metadata variable. We begin, as always, by preprocessing the data. 

```{r prep-data}

ps <- readRDS("data/cyano/phyloseq.cyano.tree.rds")
variables <- c("Microhabitat","Treatment","SampleDate")
otu <- data.frame(otu_table(ps))
tax <- data.frame(tax_table(ps)) %>% rownames_to_column("ASV")
meta <- data.frame(sample_data(ps))

timelevs <- c("Jul_2022","Nov_2022","Feb_2023", "Aug_2023","Nov_2023", "Feb_2024","Aug_2024")
treatlevs <- c("Baseline","Control","Water","N", "N+P","N+Water","P","P+Water","N+P+Water")

meta$Treatment <- factor(meta$Treatment, levels = treatlevs)
meta$SampleDate <- factor(meta$SampleDate, levels = timelevs)

```

### Shiny Dashboard

Then we enable the user to make plots as she wishes at combinations of taxonomic levels and different metadata variables. We also add in a save button to save the figures for slides, papers, proposals etc. 

```{r shiny-bubble, echo=FALSE, message=FALSE, warning=FALSE}

# ---- UI ----
all_tax_ranks <- c("Class","Order","Family","Genus")#c("Kingdom","Phylum","Class","Order","Family","Genus","Species") #Genus
all_meta_axes <- c("Treatment","SampleDate")

ui <- fillPage(
  titlePanel("Cyano Relative Abundnace Bubble Plots"),
  sidebarLayout(
    sidebarPanel(

      selectizeInput(
        "row_order_var", "Metadata variable on X axis", 
        selected = "Microhabitat", choices = all_meta_axes,
        options = list(plugins = list("remove_button"))
      ),
      
      selectizeInput(
        "tax_rank", "Taxonomic Scope",
        choices = all_tax_ranks, selected = "Order",
        options = list(plugins = list("remove_button"))
      ),
      
      

      textInput("plot_title", "Plot title", value = "Cyanobacteria Relative Abundance"),

            fluidRow(
        column(6, actionButton("refresh", "Refresh", class = "btn-primary")),
        column(6, downloadButton("save_plot", "Save figure"))
      )
    ),
    
    mainPanel(
      plotOutput("bubble_plot", width="1250px", height = "750px")
    )
  )
)

#--- SERVER ---
server <- function(input, output, session) {

  mainplot <- eventReactive(input$refresh, { #generate plot on refresh
      req(input$row_order_var)
      req(input$tax_rank)
      req(input$plot_title)
    
      ivar <- input$row_order_var
      itax <- input$tax_rank
      ititle <- input$plot_title
      
      iotu <- prop.table(as.matrix(otu), margin = 1) %>% 
        as.data.frame() %>% 
        rownames_to_column("SampleID") %>% 
        pivot_longer(!SampleID, names_to = 'ASV', values_to = "Rel_Abundance") %>% 
        left_join(., meta %>% select(SampleID, SampleDate, Microhabitat, Treatment), by='SampleID') %>% 
        left_join(., tax, by="ASV")
      
      taxord <- iotu %>% group_by(!!sym(itax)) %>% summarize(Rel_Abundance=mean(Rel_Abundance)) %>% arrange(Rel_Abundance) %>% pull(itax) 
      
      gdat <- iotu %>% group_by(!!sym(ivar), !!sym(itax), Microhabitat) %>% 
        summarize(Rel_Abundance=mean(Rel_Abundance)) %>% 
        mutate(tax = factor(!!sym(itax), levels=taxord) ) %>% arrange(tax)
      
      
      
      ggplot(gdat, aes(x = !!sym(ivar), y = tax)) +
        geom_point(aes(size = Rel_Abundance, fill = Rel_Abundance), shape = 21, color = "black", alpha = 0.9) +
        scale_size_continuous(name = "Relative Abundance", range = c(1.5, 10)) +
        scale_fill_gradient2(name = "Relative Abundance", midpoint = 0, transform = 'reverse') +
        labs(x = ivar, y = itax) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              text = element_text(size=18)) +
        facet_grid(~Microhabitat, scales = "free_x", space='free_x')+
        ggtitle(ititle)
  })
  
  output$bubble_plot <- renderPlot({
    mainplot()
  })
  
  output$save_plot <- downloadHandler(
    filename = function() {
      nm <- input$plot_title
      if (is.null(nm) || !nzchar(nm)) nm <- "bubble_plot"
      nm <- gsub("[^A-Za-z0-9_-]+", "_", nm)
      paste0(nm, ".png")
    },
    content = function(file) {
      ggsave(file, plot = mainplot(), width=5000, height = 3000, units='px')
    }
  )
  
  
}

shinyApp(ui, server)
```